Data Specialization for a Staged Lambda Calculus

### 2015-02-28 paper/relatedwork.tex
Finished pass through related work.

\paragraph{Precomputation techniques.}
Perhaps one of the most well studied example of precomputation is
partial evaluation.  Partial evaluation distinguishes between a static
(compile-time) and dynamic (run-time) stage. Given a program and the
values of static inputs (inputs that are available at compilation
time), partial evaluation generates a specialized program by
performing the computations that depend on the statics
inputs~\cite{jones96}.  The roots of partial evaluation go back to
Kleene, who in 1952 proved that a multivariate program (function)
$f(x_1, \ldots, x_{n+m})$ can be specialized for any given $a_1 \ldots
a_m$ to obtain another program $g(x_1, \ldots, x_n)$ such that $f(a_1,
\ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots, y_n)$~\cite{Kleene52}.
Kleene did not intend to improve efficiency, however.  Lombardi
appears to be the first person to use the term in his work on
incremental computation~\cite{Lombardi67}.  Since then, there has a
lot of work on partial evaluation, only broad-strokes of which we can
mention in this here.  We refer the interested reader to the excellent
book by Jones, Gommard, and Sestoft for a comprehesive discussion of
work until early 90's~\cite{JGS93}.

Earlier approaches to partial evaluation can be viewed as operating in
two stages: binding time analysis and program specialization.  For a
multivariate program with clearly marked static and dynamic arguments,
binding-time analysis identifies all the parts of the program that can
be computed by the knowledge of static arguments. Using this
information and the static arguments themselves, program
specialization specializes the program to a partially evaluated
program that can be used with many different dynamic arguments.  This
approach has been applied to construct partial evaluators for a number
of languages such as Scheme~\cite{OB91-Similix,Consel88-Schism},
considering for the most part first-order subsets.

Experience with binding time analysis showed that it can be difficult
to control, leading to programs whose performance can be difficult to
predict. This led to investigations based on type systems for making
explicit in the program the stage of each
computation~\cite{GJ91-lambda,NN92-twolevel} and writing ``meta
programs'' that when evaluated at a stage, give another program to be
evaluated at the next stage. Jorring and Scherlis describe this idea
meta programming in their paper (~\cite{JS86-staging}) as one of the
three possible approaches to taking advantage of
staging. Davies~\cite{davies96} presented a logical construction of
binding-time type systems by deriving a type system via the
Curry-Howard isomorphism applied to temporal logic.  Davies and
Pfenning proposed a new type system for staged computation based on a
particular fragment of modal logic~\cite{DP01-modal}. The work on
MetaML extended type-based techniques to a full-scale language by
developing statically typed programming language based on ML that
enables the programmer to express programs with multiple
stages~\cite{Taha97,taha-thesis-99}.  MetaML's type system is similar
to Davies~\cite{davies96} but extends it in several important ways.
Nanevksi and Pfenning further extended the these techniques by
allowing free variable to occur within staged
computations~\cite{NP05-nn}.

The type-system of \lang\ is closely related to this later line of
work on metaprogramming and staged computation.  The specific
extension to the typed lambda calculus that we use here is based on
the circle-modality of Davies~\cite{DP01-modal}.  Our types differ in
small ways to account for the specific evaluation model that we
propose.  The key difference between our work and this prior work is
our focus on the problem of pass separation: we developed an algorithm
for splitting terms in the proposed staged language.
%the staging or the splitting algorithm, which given a two-staged
%program splits it into two stages that can be executed sequentially,
%passing information from the earlier to the latter stage by
%construction.
%
Unlike partial evaluation, splitting requires no knowledge of program
argument values, only their staging.
%Using just the staging
%information, the splitting algorithm can generate programs that can
%perform non-trivial transformations on the code automatically.  As our
%examples illustrate such staging transformations can improve the
%efficiency of the source program, sometimes in asymptotically
%significant ways.
We are not aware of prior work in the context of
higher-order functional languages that presents such a splitting
algorithm.

An important class of pre-computation techniques include incremental
computation, where a program can respond to small changes to its input
efficiently by recomputing only the parts of the computation affected
by the
changes~\cite{DemersReTe81,PughTe89,RamalingamRe93,AcarBlBlHaTa09}.
Like other pre-computation techniques, incremental computation can be
thought of as a staged model, where changes (in inputs) come over a
period of time in stages.
%
However, incremental computation does not fix any of its inputs or anticipate the future
inputs, allowing in the most general case for all data to change. As a
result, its benefits are dependent on the data instance (nature of the
inputs). In addition, incremental computation techniques typically
utilize significant run-time (dynamic) information, because no
assumptions about input instances are made.  For example, we can apply
incremental computation to our quickselect example
(ref{sec:overview}), but we would not be able to achieve the bounds
that we did because incremental computation techniques would unfold
the quickselect function based on the demanded ranks, which can
require linear time (as opposed to logarithmic) at each stage.  In
addition, such an implementation would maintain rather sophisticated
data structures dynamically at run time, the benefits of which depend
highly on the forms of changes applied.

\paragraph{Frequency-reduction via pass separation.}
Perhaps the closest work to ours is the algorithm for ``data
specialization'' given by Knoblock and Ruf~\cite{knoblock96}.  Like
our work, Knoblock and Ruf are concerned about the problem of
statically splitting an explicitly staged program into two stages.
However, they only consider simple first-order language and
straight-line programs and also treat all computations guarded by
second-stage conditionals as second-stage computations.

Automatic pass seperation has also been a topic of interest in
programming systems for computer graphics, where, to achieve
high-performance, programs are typically manually separated into
components by frequency of execution (Seperation into a pre-determined
set of per-frequency modules is \emph{required} by the execution model
of modern graphics architectures.)  The software engineering
challenges modifying multiple per-stage programs have led to
suggestions of writing graphics programs in an explicitly staged
programming language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and
deferring the task of pass seperation to the compiler. However, all
prior efforts, like that of Knock and Ruf\,\cite{knoblock96} have been
limited to simple, imperative languages.

>>>>

\paragraph{Precomputation techniques.}
Perhaps one of the most well studied example of precomputation is
partial evaluation.  Partial evaluation distinguishes between static
(compile-time) and dynamic (run-time) stages. Given a program and values 
for all static inputs (inputs available at compilation
time), partial evaluation generates a specialized program by
performing computations that depend only on the static
inputs~\cite{jones96}.  The roots of partial evaluation go back to
Kleene, who in 1952 proved that a multivariate program (function)
$f(x_1, \ldots, x_{n+m})$ can be specialized for any given $a_1 \ldots
a_m$ to obtain another program $g(x_1, \ldots, x_n)$ such that $f(a_1,
\ldots, a_m, y_1, \ldots, y_n) = g(y_1, \ldots, y_n)$~\cite{Kleene52}.
Kleene did not intend to improve efficiency, however.  Lombardi
appears to be the first person to use the term in his work on
incremental computation~\cite{Lombardi67}.  There has since been 
an enormous body of work on partial evaluation, and we refer the reader to the
book by Jones, Gommard, and Sestoft for a comprehensive discussion of
work until the early 90's~\cite{JGS93}.

Early approaches to partial evaluation can be viewed as operating in
two stages: binding time analysis and program specialization.  For a
multivariate program with clearly marked static and dynamic arguments,
binding-time analysis identifies all the parts of the program that can
be computed by the knowledge of static arguments. Using this
information and the values of static arguments, program
specialization specializes the original program to a partially evaluated
one that operates on many different dynamic arguments.  This
approach has been applied to construct partial evaluators for a number
of languages such as Scheme~\cite{OB91-Similix,Consel88-Schism},
considering for the most part first-order subsets.

Experience with binding time analysis showed that it can be difficult
to control, leading to programs whose performance was difficult to
predict. This led to investigations based on type systems for making
the stage of computations explicit in programs~\cite{GJ91-lambda,NN92-twolevel}
and writing ``meta programs'' that, when evaluated at a stage, yield
a new program to be evaluated at the next stage.
%J{\o}rring and
%Scherlis describe the idea of meta programming as one of the three possible approaches to
%taking advantage of staging~\cite{JS86-staging}.
Davies~\cite{davies96} presented a
logical construction of binding-time type systems by deriving a type
system via the Curry-Howard isomorphism applied to temporal logic.
Davies and Pfenning proposed a new type system for staged computation
based on a particular fragment of modal logic~\cite{DP01-modal}. The
work on MetaML extended type-based techniques to a full-scale language
by developing statically typed programming language based on ML that
enables the programmer to express programs with multiple
stages~\cite{Taha97,taha-thesis-99}.  MetaML's type system is similar
to Davies~\cite{davies96} but extends it in several important ways.
Nanevksi and Pfenning further extended the these techniques by
allowing free variables to occur within staged
computations~\cite{NP05-nn}.

The type-system of \lang\ is closely related to this later line of
work on metaprogramming and staged computation.  The specific
extension to the typed lambda calculus that we use here is based on
the circle-modality of Davies~\cite{DP01-modal}.  Our types differ in
small ways to account for the specific evaluation model that we
propose, however the key difference between our work and this prior work is
that we instead focus on the problem of pass separation.
%we developed an algorithm
%for splitting terms in the proposed staged language.
%the staging or the splitting algorithm, which given a two-staged
%program splits it into two stages that can be executed sequentially,
%passing information from the earlier to the latter stage by
%construction.
%
%Unlike partial evaluation, splitting requires no knowledge of program
%argument values, only their staging.
%Using just the staging
%information, the splitting algorithm can generate programs that can
%perform non-trivial transformations on the code automatically.  As our
%examples illustrate such staging transformations can improve the
%efficiency of the source program, sometimes in asymptotically
%significant ways.

Another class of pre-computation techniques is incremental
computation, where a program can efficiently respond to small changes
to its input by only recomputing parts of the computation affected
by the
changes~\cite{DemersReTe81,PughTe89,RamalingamRe93,AcarBlBlHaTa09}.
Like other pre-computation techniques, incremental computation can be
thought of as a staged model, where changes in inputs come over a
period of time in stages.
%
However, incremental computation does not requiring fixing any of its
inputs and allows, in the general case, for all inputs to
change. Thus, its benefits are dependent the input changes made and
techniques typically utilize significant run-time (dynamic)
information.  For example, while we can apply incremental computation
to our quickselect example (\ref{sec:overview}), doing so would not
realize the same bounds as splitting since incremental computation
techniques would unfold the quickselect function based on the demanded
ranks, which can require linear time (as opposed to logarithmic) at
each stage.  Such an implementation would also maintain sophisticated
data structures dynamically at run time, the benefits of which depend
highly on input changes applied.

\paragraph{Frequency-reduction via pass separation.}
Perhaps the closest work to ours is the algorithm for ``data
specialization'' given by Knoblock and Ruf~\cite{knoblock96}.  Like
our work, Knoblock and Ruf are concerned about the problem of
statically splitting an explicitly staged program into two stages.
However, they only consider a simple first-order language and
straight-line programs. Their work also treats all computations
guarded by second-stage conditionals as second-stage computations,
which would prevent optimization of programs such as quickselect.

Automatic pass separation has also been a topic of interest in
programming systems for computer graphics, where, to achieve
high-performance, programs are typically manually separated into
components by frequency of execution (Separation into a pre-determined
set of per-frequency modules is \emph{required} by the execution model
of modern graphics architectures.)  The software engineering
challenges modifying multiple per-stage programs have led to
suggestions of writing graphics programs in an explicitly staged
programming language\,\cite{Proudfoot:2001,Foley:2011,He:2014}, and
deferring the task of pass separation to the compiler. However, all
prior efforts, like that of Knock and Ruf\,\cite{knoblock96} have been
limited to simple, imperative languages. We are not aware of prior
work the presents a splitting algorithm in the context of higher-order
functional languages.
